Efficient Feature Screening for Lasso-Type Problems via Batched Safe-Strong Rules

### 2020-02-17 main.tex
Add section 4.2

In fact, the solution of the primal problem (1) and the solution of the dual problem (7) can be linked by:
\begin{equation}
    \hat{\theta}(\lambda)=\frac{y-X\hat{\beta}(\lambda)}{n\lambda}=\frac{r(\lambda)}{n\lambda},
\end{equation}
so bounding $\hat{\theta}(\lambda)$ is essentially bounding the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP)\cite{wang2013lasso}, were proposed and under the standardization (3) can be simplified in the following theorems\cite{zeng2017efficient}:

\begin{theorem}[BEDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$ and $x_*=argmax_{x_j}|x_j^Ty/n|$. Under (3) and for any $\lambda\in(0,\lambda_0]$, we have $\hat{\beta}_j(\lambda)=0$ if
    \begin{equation}
        |(\lambda_0+\lambda)x_j^Ty-(\lambda_0-\lambda)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda-(\lambda_0-\lambda)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
\end{theorem}

\begin{theorem}[SEDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (3)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then $\hat{\beta}_j(\lambda_{k+1})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+1}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+1})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+1}-(\lambda_k-\lambda_{k+1})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, SEDPP rule reduces to BEDPP rule. $\hat{\beta}_j(\lambda_1)=0$ if (9) holds with $\lambda=\lambda_1$.
    \end{enumerate}
\end{theorem}

@@

\begin{theorem}[Batched EDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (3)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in the batch $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (11) or (12) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the batch and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that batch. These two computations take $O(np)$ time and the rest computation in (11) and (12) take at most a lower order of $O(p)$ time.

@@

@@

\begin{equation}
    \hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}-\frac{1}{n}\sum_{i=1}^n\{y_i\log\pi_i(\beta)+(1-y_i)\log(1-\pi_i(\beta))\}+\lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta$ is the linear function. $X$ are still the $n\times p$ standardized feature matrix  but in this model the $y\in\{0,1\}^n$ is a binary response vector. $\beta$ is still the $p\times1$ coefficient vector but we also need a $\beta_0$ as the intercept.

@@

For safe rule screening, instead of EDPP rules, Slores (sparse logistic regression screening rule)\cite{wang2014safe} is used. It was proposed as an one-step rule but we will extend it to a batched rule. Let $\Tilde{y}=2y-1\in\{-1,1\}^n$, $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$

@@

Under condition (3), for any $k=1,2,...,K-1$, assume $\hat{\beta}(\lambda_k)$, $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$, $\theta(\lambda_k)=\{r_i(\lambda_k)\Tilde{y}_i\}_{i=1}^n$ and $\hat{y}(\lambda_k):=\pi(\hat{\beta}(\lambda_k))$ are known. For some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, let

@@

\begin{itemize}
    \item Logistic regression with Slores instead of EDPP
    \item group lasso ?
    \item elastic net ?
\end{itemize}

>>>>

In fact, the solution of the primal problem (1) and the solution of the dual problem (7) can be linked by:
\begin{equation}
    \hat{\theta}(\lambda)=\frac{y-X\hat{\beta}(\lambda)}{n\lambda}=\frac{r(\lambda)}{n\lambda},
\end{equation}
so bounding $\hat{\theta}(\lambda)$ is essentially bounding the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP)\cite{wang2013lasso}, were proposed and under the standardization (2) can be simplified in the following theorems\cite{zeng2017efficient}:

\begin{theorem}[BEDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$ and $x_*=argmax_{x_j}|x_j^Ty/n|$. Under (2) and for any $\lambda\in(0,\lambda_0]$, we have $\hat{\beta}_j(\lambda)=0$ if
    \begin{equation}
        |(\lambda_0+\lambda)x_j^Ty-(\lambda_0-\lambda)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda-(\lambda_0-\lambda)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
\end{theorem}

\begin{theorem}[SEDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (2)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then $\hat{\beta}_j(\lambda_{k+1})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+1}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+1})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+1}-(\lambda_k-\lambda_{k+1})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, SEDPP rule reduces to BEDPP rule. $\hat{\beta}_j(\lambda_1)=0$ if (9) holds with $\lambda=\lambda_1$.
    \end{enumerate}
\end{theorem}

@@

\begin{theorem}[Batched EDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (2)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in the batch $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (11) or (12) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$, $x_j^Tx_*$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ and $x_j^Tx_*$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the batch and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that batch. These two computations take $O(np)$ time and the rest computation in (11) and (12) take at most a lower order of $O(p)$ time.

@@

The structure of the batched safe strong rule is easily extendable to any lasso problems that have corresponding sequential strong rule and sequential safe rule. We are going to derive the extensions to two lasso type problems and show these extensions improve computation efficiency.

@@

\begin{equation}
    \hat{\beta}(\lambda)=\underset{(\beta,\beta_0)\in \mathbb{R}^p\times\mathbb{R}^1}{\mathrm{argmin}}-\frac{1}{n}\sum_{i=1}^n\{y_i\log\pi_i(\beta)+(1-y_i)\log(1-\pi_i(\beta))\}+\lambda||\beta||_1,
\end{equation}
where $\pi(\beta)=\{\pi_i(\beta)\}_{i=1}^n=\{p(y_i=1|x_{i1},x_{i2},...,x_{ip})\}_{i=1}^n=\{e^{\eta_i(\beta)}/(1+e^{\eta_i(\beta)})\}_{i=1}^n$ is the predicted value vector and $\eta(\beta)=X\beta$ is the linear function. $X$ are still the $n\times p$ standardized feature matrix but in this model the $y\in\{0,1\}^n$ is an unstandardized binary response vector. $\beta$ is still the $p\times1$ coefficient vector but we also need a $\beta_0$ as the intercept.

@@

For safe rule screening, instead of EDPP rules, Slores (sparse logistic regression screening rule)\cite{wang2014safe} is used. It was proposed as a one-step rule but we will extend it to a batched rule. Let $\Tilde{y}=2y-1\in\{-1,1\}^n$, $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$.

@@

Under condition (2) except for standardization for $y$, for any $k=1,2,...,K-1$, assume $\hat{\beta}(\lambda_k)$, $r(\lambda_k)=y-\pi(\hat{\beta}(\lambda_k))$, $\theta(\lambda_k)=\{r_i(\lambda_k)\Tilde{y}_i\}_{i=1}^n$ and $\hat{y}(\lambda_k):=\pi(\hat{\beta}(\lambda_k))$ are known. For some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, let

@@

\subsection{Batched safe strong rule for group lasso}

The group lasso problem \cite{yuan2006model} can be defined as:

\begin{equation}
    \hat{\beta}(\lambda) = \underset{\beta\in \mathbb{R}^p}{\mathrm{argmin}}\frac{1}{2n}\bigg|\bigg|y-\sum_{g=1}^GX_g\beta_g\bigg|\bigg|_2^2+\lambda\sum_{g=1}^G\sqrt{p_g}||\beta_g||_1,
\end{equation}
where $\beta=(\beta_1^T,...,\beta_G^T)^T$, $X_g$ is a $n\times p_g$ sub-matrix consisting of columns of $X$ corresponding to features in group $g$, $g=1,2,...,G$. Besides standardization in (2), we apply an addition level of standardization at the group level\cite{breheny2015group}:

\begin{equation}
    X_g^TX_g=nI_{p_g\times p_g},\qquad g=1,2,...,G.
\end{equation}

According to the KKT condition, we can use SSR to do strong rule screening. Given $r(\lambda_k)=y-\sum_{g=1}^GX_g\hat{\beta}_g(\lambda_k)$, we discard features in the $g$-th group at $\lambda_{k+1}$ if:

\begin{equation}
    \bigg|\bigg|\frac{1}{n}X_g^Tr(\lambda_k)\bigg|\bigg|_2<\sqrt{p_g}(2\lambda_{k+1}-\lambda_k).
\end{equation}

For safe rule screening, we have SEDPP for group lasso\cite{wang2013lasso} and we can extend it the batched EDPP rule for group lasso.

\begin{theorem}[Batched EDPP for group lasso]
For the group lasso problem (22), let $\lambda_0:=\max_j\frac{||X_g^Ty||_2}{n\sqrt{p_g}}$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (2) and (23)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=\sum_{g=1}^GX_g\hat{\beta}_g(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_g(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|\left|2\lambda_{k+b}X_g^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( X_g^Ty-\frac{y^T\hat{y}(\lambda_k)X_g^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\right|_2\\
                &<2n\lambda_k\lambda_{k+b}\sqrt{p_g}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, let $X_*=argmax_{X_g}\frac{||X_g^Ty||_2}{n\sqrt{p_g}}$ with dimension $n\times p_*$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_g(\lambda_{k+b})=0$ if
        \begin{equation}
        \left|\left|(\lambda_0+\lambda_b)X_g^Ty-(\lambda_0-\lambda_b)X_g^TX_*X_*^Ty\right|\right|_2<2n\lambda_0\lambda_b\sqrt{p_g}-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2p_*}.
    \end{equation}
    \end{enumerate}
\end{theorem}

Similar to Batched EDPP rule, three expensive computations are $\{X_g^Ty\}_{g=1}^G$, $\{X_g^TX_*\}_{g=1}^G$ and $\{X_g^Tr(\lambda_k)\}_{g=1}^G$. The first and second only need to be computed once for the whole solution path. The last only needs to be computed once per batch with cost $O(np)$.
